COFFEE: an Optimizing Compiler for Finite Element Local Assembly

### 2014-06-28 TACO-paper/coffee.tex
edits

\begin{abstract}
The numerical solution of partial differential equations using the
finite element method is one of the key applications of high
performance computing. Local assembly is its characteristic
operation. This entails the execution of a problem-specific kernel for
each element in the discretized problem domain. Since the domain size
can be huge, executing efficient kernels is fundamental. Their
optimization is, however, a challenging issue. Even though affine loop
nests are generally present, the short trip counts and the complexity
of mathematical expressions make it hard to determine a single or
unique sequence of successful transformations. Therefore, we present
the design and systematic evaluation of COFFEE, a domain-specific
compiler for local assembly kernels. COFFEE manipulates abstract
syntax trees by introducing domain-aware composable optimizations aimed at
improving instruction-level parallelism, especially SIMD
vectorization, and register locality. It then generates C code
including vector intrinsics. Experiments using a range of
finite-element forms of increasing complexity show that significant
performance improvement is achieved.
\end{abstract}

@@

\section{Introduction}
In many fields, such as computational fluid dynamics, computational
electromagnetics, and structural mechanics, phenomena are modelled by
partial differential equations (PDEs). Numerical techniques, like the
finite volume method and the finite element method, are widely
employed to approximate solutions of these PDEs. Unstructured meshes
are often used to discretize the equation domain, since their
geometric flexibility allows solvers to be extremely effective. The
solution is sought by applying suitable numerical operations, or
kernels, to the components of a mesh, such as edges, vertices, or
cells. On standard clusters of multi-cores, typically, a kernel is
executed sequentially by a thread, while parallelism is achieved
partitioning the mesh and assigning each partition to a different node
or thread. Such execution model, with minor variations, is adopted,
for example, in \cite{pyop2isc}, \cite{Fenics},
\cite{fluidity_manual_v4}, \cite{lizst}.

@@

The time required to apply the numerical kernels is a major issue,
since the equation domain needs to be discretized into an extremely
large number of cells, possibly order of trillions
\cite{Rossinelli2013}, in order to obtain a satisfactory approximation
of the PDE. For example, it has been well established that mesh
resolution is critical in the accuracy of numerical weather
forecasts. However, operational forecast centres have a strict time
limit in which to produce a forecast - 60 minutes in the case of the
UK Met Office - so producing efficient kernels has a direct scientific
payoff in higher resolution, and therefore more accurate,
forecasts. Hence, the computational cost is a dominant problem in
computational science simulations, especially for those based on
finite elements, which are the subject of this paper. We address, in
particular, the well-known problem of optimizing the local assembly
phase of the finite element
method~\cite{Francis},~\cite{quadrature1},~\cite{petsc-integration-gpu},~\cite{Kirby-FEM-opt},
which can cover a significant fraction of the overall computation
run-time, often in the range 30-60$\%$. With respect to these studies,
we propose a novel set of composable code transformations targeting,
for the first time, instruction-level parallelism - with emphasis on
SIMD vectorization - and register locality.

>>>>

\begin{abstract}
The numerical solution of partial differential equations using the
finite element method is one of the key applications of high
performance computing. Local assembly is its characteristic
operation. This entails the execution of a problem-specific kernel for
each element in the discretized problem domain. Since the domain size
can be huge, executing efficient kernels is fundamental. Their
optimization is, however, a challenging issue. Even though affine loop
nests are generally present, the short trip counts and the complexity
of mathematical expressions make it hard to determine a single or
unique sequence of successful transformations. Therefore, we present
the design and systematic evaluation of COFFEE, a domain-specific
compiler for local assembly kernels. COFFEE manipulates abstract
syntax trees generated from a high-level domain-specific language for
PDEs by introducing domain-aware composable optimizations aimed at
improving instruction-level parallelism, especially SIMD
vectorization, and register locality. It then generates C code
including vector intrinsics. Experiments using a range of
finite-element forms of increasing complexity show that significant
performance improvement is achieved.
%
%The code is generated from a high level language for PDEs. An
%invocation of the kernel typically does tensor-like calculators over
%relatively small matrices; e.g., triply nested loops operating on
%about a dozen iterations. Because of the DSL context, COFFEE has a lot
%of flexibility on how it organizes/orders the calculations to tailor
%execution to a target architecture.
%
\end{abstract}

@@

\section{Introduction}
In many fields, such as computational fluid dynamics, computational
electromagnetics, and structural mechanics, phenomena are modelled by
partial differential equations (PDEs). Numerical techniques, like the
finite volume method and the finite element method, are widely
employed to approximate solutions of these PDEs. Unstructured meshes
are often used to discretize the equation domain, since their
geometric flexibility allows solvers to be extremely effective. The
solution is sought by applying suitable numerical operations, or
kernels, to the components of a mesh, such as edges, vertices, or
cells. On standard clusters of multi-cores, typically, a kernel is
executed sequentially by a thread, while parallelism is achieved
partitioning the mesh and assigning each partition to a different node
or thread. Such an execution model, with minor variations, is adopted,
for example, in \cite{pyop2isc}, \cite{Fenics},
\cite{fluidity_manual_v4}, \cite{lizst}.

@@

The time required to apply the numerical kernels is a major issue,
since the equation domain needs to be discretized into an extremely
large number of cells, possibly of the order of trillions
\cite{Rossinelli2013}, in order to obtain a satisfactory approximation
of the PDE. For example, it has been well established that mesh
resolution is critical in the accuracy of numerical weather
forecasts. However, operational forecast centres have a strict time
limit in which to produce a forecast - 60 minutes in the case of the
UK Met Office - so producing efficient kernels has a direct scientific
payoff in higher resolution, and therefore more accurate,
forecasts. Hence, the computational cost is a dominant problem in
computational science simulations, especially for those based on
finite elements, which are the subject of this paper. We address, in
particular, the well-known problem of optimizing the local assembly
phase of the finite element method
\cite{Francis,quadrature1,petsc-integration-gpu,Kirby-FEM-opt}, which
can cover a significant fraction of the overall computation run-time,
often in the range 30-60$\%$. With respect to these studies, we
propose a novel set of composable code transformations targeting, for
the first time, instruction-level parallelism - with emphasis on SIMD
vectorization - and register locality.
